AIRS: Amortized Inference for Rejection Samplers for Fast Simulator Inference

### 2019-08-20 paper/learning/notes.tex
updated - to add nested stuff tomorrow morning

  As useful rejection sampling is, it has many short comings. 
In particular, to simulate proposals for certain distributions,
such as truncated normal distributions, regardless of the easy to sample from proposal
chosen the computational running time of such rejection statements can be shown to have an
expectation equal to infinity. \textbf{Bradley: Is this true?}

Thus, the ability to learn and replace such rejection
sampling blocks with learnt surrogate functions, that are expressive enough to 
reciprocate the behavior of the rejection sampling block buys the user two things. 
A) It reduces the computational complexity of the simulator. 
B) As we now have a direct functional form of the density, we can correctly leverage
more computationally and sample efficient inference schemes.

@@

We do this by replacing the rejection sampling blocks with 
learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
This then enables us to replace rejection sampling loops with amortized functions that 
are quick to evaluate, provide a functional form to the density and significantly reduce 
memory over-heads, as compared to~\cite{baydin2018efficient}, due to not having to track the 
thousands of raw sample calls $f_{a_i}$ generated in standard simulators.

Consider the following program, that is representative of a simple stochastic simulator
and two types of rejection sampler:

\noindent
\begin{tabular}{|p{3.6cm}|p{3.6cm}|p{3.6cm}|p{3.6cm}|}
\hline
Stochastic Simulator  &  R1 & R2 & R3 \\
\hline
\begin{lstlisting}[mathescape=true]
    def f():
      $z_{1} \sim \mathcal{N}(0,1)$
      $z_{2} \gets R_{1}(z_1, 1)$
      $z_{3} \sim \mathcal{U}(0,2)$
      $z_{4} \gets R_{2}(z_2,z_3, 1)$
      $z_{5} \sim \mathcal{U}(50,30^2)$
      $z_{6} \sim R_{3}(z_5, 1)$
    return $g(z_{1:6})$
\end{lstlisting}&
\begin{lstlisting}[mathescape=true]
def $R_1(z_1)$:
  $x \sim \mathcal{N}(z_1,1)$
  if $x > 0$:
    return $x$
  else:
    return $R_{1}(z_1)$
\end{lstlisting}&
\begin{lstlisting}[mathescape=true]
def $R_2(z_2, z_3)$:
  $y \gets 0$
  while $y < z_2$:
    $y \sim \mathcal{N}(z_3,1)$
  end
  return $y$
\end{lstlisting}&
\begin{lstlisting}[mathescape=true]
def $R_3(z_5)$:
  $f(x) = \mathcal{N}(30,10^2) +$
  $\mathcal{N}(80,20^2)$
  $g(x) = \mathcal{N}(50,30^2)$
  $u \sim \mathcal{U}(0,M*g(z_5))$
  if $u \leq f(z_5)$:
    return $z_5$
  else:
    return $R_3(z_5)$
\end{lstlisting}\\
\hline
$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ & $q(z_5)$ \\
\hline
\end{tabular}

@@

If we have direct access to a rejection sampling loop then it is indeed 
possible to construct the density, up to a constant of proportionality. 
That is:
\begin{align*}
z \sim g(\theta) \\
do something \\
\alpha = p_{accept}(z,\theta) \\
\mathtt{if} Bern(\alpha) = 1: \\
return z
\end{align*}
thus the density of $y,\theta \propto q(\theta) \alpha$.
However, in order to do construct this we need the input $\theta$, $y$, $\alpha$ and the distributions utilised for the rejection sampling loop. 
For any rejection sampling block, written in a generic code base, not in the form of R3, this is an incredibly hard 
task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template. 
\textbf{Add examples in a supplement.  Need to demonstrate to YW that this is challenging to do, even if you know where 
the rejection sampling loops are. Hypothesis, RS loops are not written in a templated form}.
y leveraging the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however, 
extract all variables, other than $\alpha$ in an automated way.
However, without being able to easily extract $\alpha$ this means that we are unable to 
correctly build the form of the density and thus are limited in our ability to leverage 
powerful inference methods such as MCMC and Variational inference.
Thus, we propose another approach, automating the amortization of rejection sampling loops of type R1, R2 and R3. 
This enables us to learn functions that we can easily sample from and via the amortization process construct the density of the entire
simulator.

@@

We start by noting that the prior, defined in the context of the outer samples is defined as: 
\begin{equation}
  p(x) = 
  \begin{cases}
    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) if \mathcal{B}(x) = 1 \\
    0
  \end{cases}
\end{equation}
next, we state for a given program $\mathcal{P}$ we denote the proposal for the program as:
\begin{equation}
  q(\mathbf{x}; \kappa) = 
  \begin{cases}
  \prod^{n_x}_{j=1, j\notin S_{r}} f_{a_j}(x_j ; \psi_j)(\prod^{n_x}_{j=1, j \in S_{r}} \gamma_{a_j}(x_j; \eta_{a_j}, \psi_{j})) \texttt{ if } \mathcal{B}(x_{1:n_{x}} = 1)\\
  0
  \end{cases}
\end{equation}
where $\kappa = \{\eta_{a_j} ; a_j \in S_{r}\}$ are the variational parameters.
As we now have the density of the proposal and the density of the program we can construct a variational objective as follows:

@@

\section{Experiments: Amortizing Stochastic Processes}

\paragraph{TODO list}
\begin{itemize}
  \item Implement baselines
  \item Implement plotting scripts and evaluation scripts
  \item testing code and infrastructure
  \item litrature review on any methods to learn rejection sampling methods, 
  and conditional density estimation. See this paper for kernel approximation 
  methods~\cite{erraqabi2016pliable}. In addition to the methods listed in the 
  rejection sampling subsection. 
\end{itemize}

@@

\subsection{Baseline}
As a first approach, we train a regressor to learn a function to map
between the input / output pairs that are outputted from the rejection
sampling blocks with in the simulator.

@@

\section{Friday morning}
The subprograms within the simulator should be treated as nested probabilistic programs.

@@

\subsection{Method 1: Learn the}
$\gamma_{a_{j}}(\cdot ; \eta_{a_{j}}, \cdot)$

@@

The things inside of the rejection sampler are proportional to the following:
% \begin{equation}
%   \begin{cases}
%   \end{cases}
% \end{equation}

@@

If you minimise an $L_2$ norm the optimal 
\section{ Questions for Tom}

>>>>

However, in many of the non-standard ABC forms of the rejection samplers, the acceptance 
probability is completely intractable, or is very difficult to calculate. 
Thus, as useful rejection sampling is for generating samples from densities that we 
would not be able to sample from, it has many short comings. 
These are of course well known and it is the reason for developing more sophisticated 
sampling schemes, such as importance sampling based methods.
However, in the setting of such complex simulators, written over many years and
in convoluted code bases, there is no easy way to re-implement such sampling 
schemes.
Hence, to leverage existing frameworks we can make use of the probabilistic 
programming executation (PPX) protocols~\cite{baydin2018efficient}. 
These protocols allow one to exact all raw sample calls $f_{a_j}$ from the
simulator. This then enables one to easily construct a prior for the full program
$P(\mathcal{T} = x_{1:n_{x}}| \lambda)$. 
In addition to this, once such rejection schemes, \emph{nested programs}, are located, we have a number of 
primatives that we can leverage to isolate calls to those addresses during training, to collect the samples generated during 
\texttt{.forward()} runs of the simulator, 
and during the inference procedure replace calls to those addresses $a_j$ with a
learnt proposal $\gamma_{a_j}(x_j ; \eta_{a_j}, \phi_j ) $.

% In particular, to simulate proposals for certain distributions,
% such as truncated normal distributions, it can be very difficult to
% hand-craft a proposal such that the acceptance rate.

Thus, the ability to learn and replace such rejection
sampling blocks with learnt surrogate functions, that are expressive enough to 
reciprocate the behavior of the rejection sampling block provides the user with three things. 
\begin{itemize}
  \item It reduces the computational complexity of the simulator.  
  \item As we now have a direct functional form of the density, we can correctly leverage more computationally and sample efficient inference schemes. 
  \item By only amortizing sub-programs and not amortizing the whole simulator, the larger structure of the simulator remains interpretable, in the sense that we can 
  still decode the decisions that it is making.  
\end{itemize}

@@

% We do this by replacing the rejection sampling blocks with 
% learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
% This then enables us to replace rejection sampling loops with amortized functions that 
% are quick to evaluate, provide a functional form to the density and significantly reduce 
% memory over-heads, as compared to~\cite{baydin2018efficient}, due to not having to track the 
% thousands of raw sample calls $f_{a_i}$ generated in standard simulators.

\section{A Stochastic Simulator}

In this section, we present a simple stochastic simulator and povide 
examples of the rejection blocks found within standard stochastic
simulators~\cite{smith2008towards,bershteyn2018implementation, gleisberg2009event}. 
This simulator will form the basis for our experiemnts in Section~\ref{sec:experiments}.
Consider the following program, that is representative of a simple stochastic simulator
and two types of rejection sampler:

\noindent
\begin{tabular}{|p{3.6cm}|p{3.6cm}|p{3.6cm}|p{3.6cm}|}

\hline
Stochastic Simulator  &  R1 & R2 & R3 \\
\hline
\begin{lstlisting}[mathescape=true]
    def s():
      $z_{1} \sim \mathcal{N}(0,1)$
      $z_{2} \gets R_{1}(z_1, 1)$
      $z_{3} \sim \mathcal{U}(0,2)$
      $z_{4} \gets R_{2}(z_2,z_3, 1)$
      $z_{5} \sim \mathcal{U}(50,30^2)$
      $z_{6} \sim R_{3}(z_5, 1)$
    return $g(z_{1:6})$
\end{lstlisting}&
\begin{lstlisting}[mathescape=true]
def $R_1(z_1)$:
  $x \sim \mathcal{N}(z_1,1)$
  if $x > 0$:
    return $x$
  else:
    return $R_{1}(z_1)$
\end{lstlisting}&
\begin{lstlisting}[mathescape=true]
def $R_2(z_2, z_3)$:
  $y \gets 0$
  while $y < z_2$:
    $y \sim \mathcal{N}(z_3,1)$
  end
  return $y$
\end{lstlisting}&
\begin{lstlisting}[mathescape=true]
def $R_3(z_5)$:
  $f(x) = \mathcal{N}(30,10^2) +$
  $\mathcal{N}(80,20^2)$
  $g(x) = \mathcal{N}(50,30^2)$
  $u \sim \mathcal{U}(0,M*g(z_5))$
  if $u \leq f(z_5)$:
    return $z_5$
  else:
    return $R_3(z_5)$
\end{lstlisting}\\
\hline
$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ & $q(z_5)$ \\
\hline
\end{tabular}

@@

If we have direct access to a rejection sampling loop then it is indeed 
possible to construct the density, up to a constant of proportionality. 
% That is:
% \begin{align*}
% z \sim g(\theta) \\
% do something \\
% \alpha = p_{accept}(z,\theta) \\
% \mathtt{if} Bern(\alpha) = 1: \\
% return z
% \end{align*}
 $x,z \propto f(\theta) p(accept)$.
However, in order to do construct this we need the input $z$, the sample from the proposal $x$ and $p(accept)$ (should use the notation above).
For any rejection sampling block, written in a generic code base, not in the form of R3, this is an incredibly hard 
task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template. 
\textbf{Add examples in a supplement.  Need to demonstrate to YW that this is challenging to do, even if you know where 
the rejection sampling loops are. Hypothesis, RS loops are not written in a templated form}.
y leveraging the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however, 
extract all variables, other than $P(accept)$ in an automated way.
As without being able to easily extract $p(accept)$ this means that we are unable to 
correctly build the form of the density and thus are limited in our ability to leverage 
powerful inference methods such as MCMC and Variational methods.
Thus, we propose another approach, automating the amortization of rejection sampling loops of type R1, R2 and R3. 
This enables us to learn functions that we can easily sample from and via the amortization process construct the density of the entire
simulator.

@@

We start by noting that the prior, defined in the context of the outer samples is defined as: 
\begin{equation}
  p(x) = 
  \begin{cases}
    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) \texttt{ if } \mathcal{B}(x) = 1 \\
    0
  \end{cases}
\end{equation}
next, we state for a given program $\mathcal{P}$ we denote the proposal for the program as:
\begin{equation}
  q(\mathbf{x}; \kappa) = 
  \begin{cases}
  \prod^{n_x}_{j=1, j\notin S_{r}} f_{a_j}(x_j ; \psi_j)(\prod^{n_x}_{j=1, j \in S_{r}} \gamma_{a_j}(x_j; \eta_{a_j}, \psi_{j})) \texttt{ if } \mathcal{B}(x_{1:n_{x}} = 1)\\
  0
  \end{cases}
\end{equation}
where $\kappa = \{\eta_{a_j} ; a_j \in S_{r}\}$ are the variational parameters.
As we now have the density of the proposal and the density of the program we can construct a variational objective as follows:

@@

\section{Experiments: Amortizing Stochastic Processes}
\label{sec:experiments}
\paragraph{TODO list}
\begin{itemize}
  \item Implement baselines
  \item Implement plotting scripts and evaluation scripts
  \item testing code and infrastructure
  \item litrature review on any methods to learn rejection sampling methods, 
  and conditional density estimation. See this paper for kernel approximation 
  methods~\cite{erraqabi2016pliable}. In addition to the methods listed in the 
  rejection sampling subsection. 
\end{itemize}

@@

\subsection{Baseline}
For gif plots see the plots folder in code$\backslash$amortized\_rs$\backslash$plots

As a first approach, we train a regressor to learn a function to map
between the input / output pairs that are outputted from the rejection
sampling blocks with in the simulator.

@@

\section{The difference between this paper and Inference compilation paper}

The main fundemental difference is that IC cares about the full global program, 
we only care about the local (sub / nested programs).

We are not interesting in calculating the weights $w^k$ since the method that we 
develop here is applicable to any inference scheme, in theory.

In the IC paper you are approximating $p(x|y)$ the \emph{the global} program, where 
as you can view this work as learning an approximation for the \emph{local}, sub-programs
$P^{in}(x|\nu)$ where the nuissince parameters in this paper are represented by $\psi$.

In the notation of the inference compilation paper we essentailly learn $q_{a_{t},i_{t}}(x_t| x_{1:t-1}, y)$
where the indicies are part of the set $S_r$.
 The difference here though, is that we only care about the input at that address, your proposals 
 are condiitoning on everything upto that point of the executation trace $\mathcal{T}$
 in the simulator. 
 The training data represented by (11) in the IC paper is not the training data we use in this paper,
 we only care about the $x^{m}_t$ or $z$'s and $x$'s in this paper (notation is not entirely consitant throuhg out this paper yet)
 But these are the parameters that are generated from raw sample calls within side the RS loop $f_{a_i \in S_r}(\cdot))$.

Thus the density of the full joint of the program in this setting is given as:
\begin{equation}
  p(x,y) = \prod_{t=1, t \notin S_r} f_{a_t}(x_t | \psi_t) \prod_{n=1} g_{b_n}(y_n | \phi) \prod_{h=1}\gamma_{a_j \in S_r}(\cdot)
\end{equation}
\section{Friday morning}
The subprograms within the simulator should be treated as nested probabilistic programs.

@@

\subsection{Method 1: Learn the maoritzation of the sub-programs}
That is learn $\gamma_{a_{j}}(\cdot ; \eta_{a_{j}}, \cdot)$ for each program in the set $S_r$.

@@

The things inside of the rejection sampler are proportional to the following:
\begin{equation}
  P_{a_j}^{in}(x_j ; \phi_j) \propto
  \begin{cases}
  \prod^{n^{z}_{a_j}} f^{in}_{i}(z_{ij} | \theta_i) \prod_{k=1}^{n^{y}_{b_k}}g^{in}_{k}(y_i ; \phi_{i}) \texttt{ if } \mathcal{B}(z_{1:n^{z}_{a_j}}) = 1 \\
  0
  \end{cases}
\end{equation}

Thus, we need a regression from $\phi_j \rightarrow P^{in}){a_j}(\phi_j) $.
We have access to pairs $\{\phi_j, \hat{P}^{in}_{a_j}(\phi_j)\}$ for which 
the expectation $\mathbb{E}[\hat{P}^{in}_{a_j}( \phi_j)| \phi_j] = P^{in}_{a_j} (\phi_j)$ where
$\hat{P}$ is \textbf{Play back convo to find out}

@@

If you minimise an $\mathcal{L}_2$ norm for this given problem then the optimal proposal natural arises, see~\ref{fig:tomfig} 
The training regime would still need to be non-bias.

\begin{figure}[h!]
  \centering
  \includegraphics[width=\textwidth]{tom_derivation.jpg}
  \caption{Tom's derivation proving that the $\mathcal{L}_2$ norm in this problem set-up enables us to learn a proposal that exactly
  matches the true distribution.  }
  \label{fig:tomfig}
\end{figure}
\section{ Questions for Tom}
